---
title: "Data cleaning and EDA"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---


```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(readr)
library(ggplot2)
library(table1)
library(cowplot)
library(broom)
library(heatmaply)
library(RColorBrewer)
```


## Data Description: 

The dataset was collected by the Behavioral Risk Factor Surveillance System (BRFSS) in 2015. This original dataset contains responses from 441,455 individuals and has 330 features. For this project, we use the one with 253,680 responses and a three-class diabetes target variable indicating no diabetes, prediabetes, or diabetes. The potential factors for diabetes are biological factors such as sex, blood pressure, cholesterol, BMI, heart disease, social factors such as education and income, and habits of smoking, physical exercise, eating fruit and vegetables.

There are 22 variables in our original dataset.

* `Diabetes_012`: This is a categorical variable indicating the diabetes condition of the individual. 0 is for no diabetes or only during pregnancy, 1 is for prediabetes, and 2 is for diabetes.

* `HighBP`: This is a categorical variable indicating whether or not an individual has high blood pressure. (0 = no, 1 = yes)

* `HighChol`: This is a categorical variable indicating whether or not an individual has high blood cholesterol. (0 = no, 1 = yes)

 * `Cholcheck`: This is a categorical variable indicating whether or not an individual has cholesterol check in 5 years. (0 = no, 1 = yes)
 
* `BMI`: It is a continuous variable indicating individual's Body Mass Index.

* `Smoker`: It is a categorical variable showing participants identify themselves whether they have smoked at least 100 cigarettes in their entire life. (0 = no, 1 = yes)

* `Stroke`: A categorical variable indicating whether the individual was told they had a stroke. (0 = no, 1 = yes)

* `HeartDiseaseorAttack`: A categorical variable showing whether the individual had coronary heart disease (CHD) or myocardial infarction (MI). (0 = no, 1 = yes)

* `PhysActivity`: A categorical variable showing whether individual has physical activity in past 30 days which are not included in job. (0 = no, 1 = yes)

* `Fruits`: A categorical variable showing whether the individual consumes fruit 1 or more times per day. (0 = no, 1 = yes)

* `Veggies`: A categorical variable showing whether the individual consumes vegetables 1 or more times per day. (0 = no, 1 = yes)

* `HvyAlcoholConsump`: A categorical variable showing whether the individual is a heavy drinker, which means adult men having more than 14 drinks per week and adult women having more than 7 drinks per week. (0 = no, 1 = yes)

* `AnyHealthcare`: A categorical variable showing whether the individual has any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc. (0 = no, 1 = yes)

* `NoDocbcCost`: A categorical variable showing whether the individual has a time in the past 12 months when he/she needed to see a doctor but could not because of cost. (0 = no, 1 = yes)

* `GenHlth`: A categorical variable showing the self evaluation of the individual's general health. (1 = excellent, 2 = very good, 3 = good, 4 = fair, 5 = poor)

* `MentHlth`: A categorical variable showing the self evaluation of the individual was showing mental health problems including stress, depression, and problems with emotions, for how many days in their past 30 days.

* `PhysHlth`: A categorical variable showing the self evaluation of the individual was showing physical health problems, which includes physical illness and injury, for how many days during the past 30.

* `DiffWalk`: A categorical variable showing the self evaluation of whether the individual has serious difficulty walking or climbing stairs. (0 = no, 1 = yes)

* `Sex`: A categorical variable showing the individual's born sex. (0 = female, 1 = male)

* `Age`: A categorical variable showing the individual's age. (1 = 18-24 9 = 60-64 13 = 80 or older)

* `Education`: A categorical variable showing the individual's Education level in the scale of 1-6. [1 = Never attended school or only kindergarten, 2 = Grades 1 through 8 (Elementary), 3 = Grades 9 through 11 (Some high school), 4 = Grade 12 or GED (High school graduate), 5 = College 1 year to 3 years (Some college or technical school), 6 = College 4 years or more (College graduate)]

* `Income`: A categorical variable showing the individual's income level in the scale of 1-8. (1 = less than 10,000 dollars, 5 = less than 35,000 dollars, 8 = 75,000 dollars or more)

To explore our question of interest, we will keep `Diabetes_012` as our dependent variable, and investigate the relationship between the other variables and `Diabetes_012`. In our hypothesis, biological factors including blood pressure, cholesterol level, BMI, Heart Disease, age, and sex might have a relationship with the levels of diabetes, so we keep variables `HighBP`, `HighChol`, `BMI`, `HeartDiseaseorAttack`, `Age`, and `Sex`. 

We also predict that social factors including income, education, smoking habits, alcohol consumption, physical exercise frequency, and fruit and vegetable eating will be associated with different diabetes levels. Therefore, we want to keep `Income`, `Education`, `Smoker`, `PhysActivity`, `Fruits`, `HvyAlcoholConsump`, and `Veggies` variables.

## Read in and Clean Data  


First, we read in our data from the csv file. 

```{r, echo = FALSE}
diabetes_raw0 <- read_csv("data/Diabetes_012_data.csv")

head(diabetes_raw0)
```



```{r}
diabetes_raw  = diabetes_raw0 |>
  dplyr::select(Diabetes_012,HighBP,HighChol, BMI, Smoker, 
         HeartDiseaseorAttack, Sex,Income, Education, 
         PhysActivity, Fruits, Veggies, Age, HvyAlcoholConsump) |>
   mutate(Diabetes_012 = factor(Diabetes_012, levels = c(0, 1, 2), labels = c("No Diabetes", "Pre-Diabetes", "Diabetes")),
         HighBP = factor(HighBP, levels = c(0, 1), labels = c("No", "Yes")),
         HighChol = factor(HighChol, levels = c(0, 1), labels = c("No", "Yes")),
         HeartDiseaseorAttack = factor(HeartDiseaseorAttack, levels = c(0, 1), labels = c("No", "Yes")),
         Smoker = factor(Smoker, levels = c(0, 1), labels = c("Nonsmoker", "Smoker")),
         Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
         Fruits = factor(Fruits, levels = c(0, 1), labels = c("No", "Yes")),
         Veggies = factor(Veggies, levels = c(0, 1), labels = c("No", "Yes")),
         HvyAlcoholConsump = factor(HvyAlcoholConsump, levels = c(0, 1), labels = c("No", "Yes")),
         PhysActivity = factor(PhysActivity, levels = c(0, 1), labels = c("No", "Yes")))

diabetes_raw
```

## Statistical Summary of Variables

This table presented a statistical summary of the variables in our dataset.

```{r}
summary(diabetes_raw)
```


## Exploratory Data Analysis

### Bivariate Plots and Analysis Analysis

We first perform univariate analysis for our dependent variable `Diabetes_012` and plot the bivariate relationship between our interested factors with the diabetes variable respectively. 


#### Distribution of Diabetes Types
```{r}
Diabetesraw_dist = 
  ggplot(diabetes_raw, aes(Diabetes_012)) + 
  geom_bar(aes(fill = Diabetes_012)) +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes Status") +
  ggtitle("Raw Overall Diabetes Status Distribution") +
  theme_minimal()

Diabetesraw_dist
```


From the distribution of Diabetes Status, we can see that the frequency of No Diabetes group is very high with more than 200,000. The Diabetes group has around 30,000 cases, and Pre-Diabetes has less than 10,000 cases. We realize that Pre-Diabetes status has extreme low data samples, so we will exclude this group of data in our future analysis.

```{r}
diabetes_new = 
  diabetes_raw |>
  filter(Diabetes_012 == 'Diabetes' | Diabetes_012 == 'No Diabetes') |>
  rename(Diabetes_status = Diabetes_012)

#diabetes_new
write.csv(diabetes_new, "data/diabetes_new.csv", row.names = FALSE)

diabetes_new1=diabetes_new|>
  sample_n(200, replace = FALSE)
write.csv(diabetes_new1, "data/diabetes_new1.csv", row.names = FALSE)
```


####  1. High Blood Pressure via Diabetes
```{r}
HighBP_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = HighBP), position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes Status") +
  ggtitle("HighBP over Diabetes") +
  theme_minimal()

HighBP_dist
```

The graph presents a comparison of the occurrence of high blood pressure in individuals with and without diabetes, showing a higher frequency of individuals without diabetes and without high blood pressure, a moderate frequency of those without diabetes but with high blood pressure, and lower frequencies for both categories among individuals with diabetes. From the graph, we can see that the high blood pressure is more prevalent in diabetes group than in no diabetes group.

####  2. High Cholesterol Level via Diabetes
```{r}
HighChol_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = HighChol), position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes Status") +
  ggtitle("HighChol over Diabetes") +
  theme_minimal()

HighChol_dist
```

From the graph, we can see that the high blood pressure is more prevalent in Diabetes group than in No Diabetes group.


####  3. BMI via Diabetes
```{r}
BMI_dist = 
  ggplot(diabetes_new, aes(BMI, ..count..)) + 
  geom_bar(aes(fill = Diabetes_status)) +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("BMI") +
  ggtitle("BMI Distribution") +
  theme_minimal() 

BMI_dist
```

For both diabetes and no diabetes group, the BMI of the population are approximately normal with a slight right tail. The BMI distribution for the diabetes group extends further to the right, indicating a presence of higher BMI values in this group compared to those without diabetes.


####  4. Heart Disease or Attack via Diabetes
```{r}
Heartdisease_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = HeartDiseaseorAttack), position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes Status") +
  ggtitle("Heart Disease over Diabetes") +
  theme_minimal()

Heartdisease_dist
```

From the graph, heart disease is in a larger proportion in the Diabetes group than in the No Diabetes group.

####  5. Age via Diabetes
```{r}
Age_dist = 
  ggplot(diabetes_new, aes(Age, ..count..)) + 
  geom_bar(aes(fill = Diabetes_status)) +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Age") +
  ggtitle("Age Distribution") +
  theme_minimal()

Age_dist
```

We can see that for both Diabetes and No Diabetes groups, more people are in their age of group 9 or group 10, which indicates they are between 60-70 years old. In younger age groups, there are less population in our dataset.

####  6. Sex via Diabetes
```{r}
Sex_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = Sex), , position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes") +
  ggtitle("Sex over Diabetes Status") +
  theme_minimal()

Sex_dist
```

There are approximately same number of females and males in Diabetes group, and there are more females than males in the No Diabetes group. 


####  7. Income via Diabetes
```{r}
Income_dist = 
  ggplot(diabetes_new, aes(Income, ..count..)) + 
  geom_bar(aes(fill = Diabetes_status)) +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Income") +
  ggtitle("Income Distribution") +
  theme_minimal()

Income_dist
```
For Diabetes group, the income distribution was relatively flat from 1 to 8. For No Diabetes group, the income distribution is highly skewed with the highest frequency in the highest income level.

####  8. Education via Diabetes
```{r}
Education_dist = 
  ggplot(diabetes_new, aes(Education, ..count..)) + 
  geom_bar(aes(fill = Diabetes_status)) +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Education") +
  ggtitle("Education Distribution") +
  theme_minimal()

Education_dist
```
In the No Diabetes group, the education is highly skewed with more frequency in high education levels, while in Diabetes group, the frequency of education level remains the same from 4 to 6. This means that they are in High school graduate, College 1 year to 3 years, and College 4 years or more education levels.


####  9. Fruits via Diabetes
```{r}
Fruits_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = Fruits), , position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes") +
  ggtitle("Fruits over Diabetes") +
  theme_minimal()

Fruits_dist
```

There is a higher proportion of fruit eating in the No Diabetes group than the Diabetes group. 



####  10. Veggies via Diabetes
```{r}
Veggies_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = Veggies), , position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes") +
  ggtitle("Vegetables over Diabetes") +
  theme_minimal()

Veggies_dist
```

For both Diabetes and No Diabetes groups, more people tend to eat veggies. 


####  11. Smoker via Diabetes
```{r}
Smoker_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = Smoker), , position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes") +
  ggtitle("Smokers over Diabetes") +
  theme_minimal()

Smoker_dist
```

More people are smokers than nonsmokers in the Diabetes group, while more people are smokers than nonsmokers in the No Diabetes group. 

####  12. PhysActivity via Diabetes
```{r}
Physactivity_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = PhysActivity), , position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes") +
  ggtitle("PhysActivity over Diabetes") +
  theme_minimal()

Physactivity_dist
```
Higher proportion of people have physical activities in the No Diabetes group comparing to the Diabetes group.


####  13. HvyAlcoholConsump via Diabetes
```{r}
HvyAlcohol_dist = 
  ggplot(diabetes_new, aes(Diabetes_status, ..count..)) + 
  geom_bar(aes(fill = HvyAlcoholConsump), , position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Diabetes") +
  ggtitle("Alcohol over Diabetes") +
  theme_minimal()

HvyAlcohol_dist
```

According to the bar chart distribution, we decide to drop Heavyalcoholconsumption, heart disease attack these two variables due to following reasons. 

1. For the heavy alcohol consumption, the samples drinking alcohol in large amount is overall too few in the sample, so the variable more not be fully accutately captured. Therefore, if we test on this variable, it may be not representitive and significant. 

2. For the heart disease attack, it is the same reason that sample size with heart disease attack is too small, so it may not be representitive and meaningful to test on. 

We will also drop age and education these two variables due to following reasons. 

1. For the age, the measure and categorize method is not suitable for our statistical test, since it is a categorical variable showing the individual's age. (1 = 18-24 9 = 60-64 13 = 80 or older), and each number does not represent the true age of the samples, so we cannot do a precise test on it. 

2. For the education, it is the same reason that it is a categorical variable showing the individual's Education level in the scale of 1-6. [1 = Never attended school or only kindergarten, 2 = Grades 1 through 8 (Elementary), 3 = Grades 9 through 11 (Some high school), 4 = Grade 12 or GED (High school graduate), 5 = College 1 year to 3 years (Some college or technical school), 6 = College 4 years or more (College graduate)].

**Therefore, now we come to close with 9 variables that are potential estimates in our model. They are: 1. High Blood Pressure. 2. High Cholesterol Level. 3. BMI. 4. Sex. 5. Fruits Consumption. 6. Veggies Consumption. 7. Smoking. 8. Income Level. 9. Physical Acitivity.**

### Plots all together
```{r}
plotsall = plot_grid(
  HighBP_dist + theme(legend.position="none"),
  HighChol_dist + theme(legend.position="none"), 
  Fruits_dist + theme(legend.position="none"), 
  Veggies_dist + theme(legend.position="none"),
  BMI_dist + theme(legend.position="none"),
  Sex_dist + theme(legend.position="none"),
  Income_dist + theme(legend.position="none"),
  Physactivity_dist + theme(legend.position="none"),
  Smoker_dist + theme(legend.position="none")) +
  theme(axis.text.x = element_text(size = 4))
plotsall

```
We now can see all the 9 variables distributions from our plot together. 

### Plots for further investigation: After we investigate all the 9 variables distribution, we are now ready to investigate further in specific variables that we are interested in.
- Firstly, we decide to make one box plot for the numeric values variables like BMI. 
- Then since veggies and fruits consumptions are two variables close related, we also need to generate a plot for these 2 variable interelationship. 
- To further visualize the variable stats, we also build a R table to investigate each factor's significance.
- Finally, we also make a small heat map to show the correlation between each variable to make sure that we are including only the variables with little correlation to each other to inhibit the collinearity problem. 


####  1. Box plot distribution of BMI in different Diabetes status
```{r}
diabetes_bmi = 
  diabetes_new |>
  filter(Diabetes_status == "Diabetes" | Diabetes_status == "No Diabetes")

plotBMI = 
        ggplot(diabetes_bmi, aes(x = BMI, y = Diabetes_status, fill = Diabetes_status))+
        geom_boxplot(alpha = 0.5) +
        theme(legend.position = 'none') +
        scale_fill_brewer(palette = 'Pastel1') +
        ylab("Diabetes Status") +
        xlab("BMI Value") +
        ggtitle("Box Plot Distribution of BMI over Diabetes Status") +
        theme_minimal()
  
plotBMI
```
From the box plot for BMI value, we can see that Diabetes group BMI has a higher median which is around 30 than the median in non diabetes group of which is around 27. The No diabetes group also has a left shifted distribution than the diabetes group meaning that generally Diabetes group has a higher BMI value distribution than the non diabetes group. 


####  2. Mosaic plot to see distribution of Fruits and Veggies
```{r}
Mosaicplot = 
  mosaicplot(Fruits ~ Veggies, data=diabetes_new,
          labeling = labeling_cells(text = round(x, 2)), 
          legend = TRUE,
          main = "Fruits and Veggies Consumption Mosaic Map",
          xlab = "Veggies Consumption",
          ylab = "Fruits Consumption",
          las = 0.5, color = brewer.pal(2, name = "Pastel1")) 

Mosaicplot
```
Our mosaic map shows that through the whole sample, a large amount of people have a good habit of consuming both fruits and veggies, and only a small number of people consume veggies without fruits and quite a few people only consume fruits. There are also a small number of people who do not consume fruits and veggies. 

#### Test for P-value function
```{r}
pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a one-side anova test
        p <- aov(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}
```


#### 3. R table 1 for 9 variables with P-value attached
```{r}
## Rtable_1 shows the demographics stats
Rtable_1 = 
  table1(~ factor(Sex) + factor(HighBP) + factor(HighChol) + BMI + factor(Smoker) + factor(Income) + factor(PhysActivity) + factor(Fruits) + factor(Veggies) | Diabetes_status, data=diabetes_new, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_1
```
From the R table 1, we can see that all the categorical variables have p-values smaller than 0.001 which means that the variables are good to test and may be significant in our further analysis, so we can include them all in. 
From the table, we can clearly see that both proportions of high cholesterol and high blood pressure are higher in diabetes than in no diabetes. Male proportions in diabetes are higher than females. Diabetes group BMI has a mean of 31.9 which is much higher than the mean in non diabetes group of 27.7. There are higher proportions of smokers in the diabetes group than in the non diabetic group. The no diabetes group has higher proportions of high level incomes than the no diabetes group. Diabetes group also generate a lower proportion of physical activity than the non diabetes group. Diabetes group also has a lower proportion of people consuming fruits (58.5% ) and veggies (75.6%) than the non-diabetes group (fruits:64.3% veggies:82.1%).



#### 4. Heat map to see correlation between each variables.
```{r}

options(repr.plot.width=30, repr.plot.height=40)

heatmapplot = 
          heatmaply(cor(diabetes_raw0), 
          col = brewer.pal(8, name = "PiYG"),
          main = "Correlation Heatmap", 
          margins = c(100, 100),
          notecol = "Black", 
          symmetric = TRUE)

heatmapplot
```

The correlation heatmap presents a visual representation of the strength and direction of the relationships between various health, lifestyle, and socioeconomic variables in relation to diabetes. The color spectrum from dark blue to bright yellow indicates the range of correlation coefficients, with darker shades representing weaker correlations and brighter shades indicating stronger ones. Notably, variables such as High Blood Pressure (HighBP), High Cholesterol (HighChol), and Body Mass Index (BMI) show brighter hues along the diagonal, suggesting a stronger positive correlation with the diabetes outcome (Diabetes_012). This implies that higher values of these factors are associated with a greater likelihood of diabetes. In contrast, variables like vegetable consumption (Veggies) exhibit a dark shade when correlated with High Cholesterol (HighChol), indicating a weak, potentially negative relationship.
