---
title: "Data cleaning and EDA"
author: "Qianying Wu"
date: "2023-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ggplot2)
library(table1)
```

## Read in Data

```{r}
diabetes_raw <- read_csv("data/Diabetes_012_data.csv") |>
  # factor Diabetes to 0, 1, 2
  mutate(Diabetes_012 = factor(Diabetes_012, levels = c(0, 1, 2)),
         HighBP = factor(HighBP, levels = c(0, 1), labels = c("No", "Yes"))) 

head(diabetes_raw)

```

## Exploratory Data Analysis

### Univariate Analysis

####  Diabetes Type

```{r}

diabetes_raw$Diabetes_012 <- factor(diabetes_raw$Diabetes_012, levels = c(0, 1, 2),
                                     labels = c("No Diabetes", "Pre-Diabetes", "Diabetes"))

# Use ggplot2 to plot the frequency chart
ggplot(data = diabetes_raw) +
  geom_bar(mapping = aes(x = Diabetes_012), fill = "skyblue") +
  labs(x = "Diabetes Status", y = "Frequency", title = "Frequency of Diabetes Categories") +
  theme_minimal()
```

####  High Blood Pressure

```{r}
ggplot(data = diabetes_raw) +
  geom_bar(mapping = aes(x = HighBP), fill = "lightgreen") +
  labs(x = "High Blood Pressure", y = "Frequency", title = "Frequency of High Blood Pressure") +
  theme_minimal()
```

#### bmi categorization

```{r}
# write a function to categorize bmi values to diverse weight status
categorize_bmi <- function(bmi) {
  if (bmi < 18.5) {
    return("Underweight")
  } else if (bmi >= 18.5 & bmi < 25) {
    return("Normal weight")
  } else if (bmi >= 25 & bmi < 30) {
    return("Overweight")
  } else {
    return("Obesity")
  }
}
# add a column named weight_status to the data set
diabetes_raw$weight_status <- sapply(diabetes_raw$BMI, categorize_bmi)

```

#### Plot: Association between Diabetes Status and blood pressure status
```{r}
ggplot(diabetes_raw, aes(x = Diabetes_012, fill = HighBP)) +
  geom_bar(position = "fill") +
  ylab("Blood Pressure Status") +
  xlab("Diabetes Status") +
  ggtitle("Association between Diabetes Status and Blood Pressure Status") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal()
```

#### Plot: Association between Diabetes Status and Weight Status
```{r}
ggplot(diabetes_raw, aes(x = Diabetes_012, fill = weight_status)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  xlab("Diabetes Status") +
  ggtitle("Association between Diabetes Status and Weight Status") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal()
```

#### Plot: Box plot distribution of BMI in different Diabetes status
```{r}
plotBMI = ggplot(diabetes_raw, aes(x = BMI, y = Diabetes_012, fill = Diabetes_012))+
        geom_boxplot(alpha = 0.5) +
        theme(legend.position = 'none') +
        scale_fill_brewer(palette = 'Pastel1') +
        ylab("Diabetes Status") +
        xlab("BMI Distribution") +
        ggtitle("Box Plot Distribution of BMI over Diabetes Status") +
        theme_minimal()
        
  
  
plotBMI
```

#### Plot: Smoking Status over Diabetes Status
```{r}
plot_smoke = 
  ggplot(diabetes_raw, aes(Smoker, ..count..)) + 
  geom_bar(aes(fill = Diabetes_012), position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Smoking Status") +
  ggtitle("Smoking Status Frequencies over Diabetes Status") +
  theme_minimal()

plot_smoke
```
### Plot: Fruits or Veggies over different Diabetes Status
```{r}
mosaicplot(Fruits~Veggies,data=diabetes_raw,col=c("Light Pink","Light Blue"))
```



#### Plot: Distribution of Education Level over different Diabetes Status
```{r}
plot_education = 
  ggplot(diabetes_raw, aes(x = Education, ..count..)) + 
  geom_bar(aes(fill = Diabetes_012), position = 'dodge') +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Education Level") +
  ggtitle("Education Level Frequencies over Diabetes Status") +
  theme_minimal()

plot_education
```

#### Plot: Distribution of Income Level over different Diabetes Status
```{r}
plot_income = 
  ggplot(diabetes_raw, aes(Income, ..count..)) + 
  geom_bar(aes(fill = Diabetes_012), position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Income Level") +
  ggtitle("Income Level Frequencies over Diabetes Status") +
  theme_minimal()

plot_income
```

#### Statistical Test 1.1 P-value function
```{r}
pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a one-side anova test
        p <- aov(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}
```


#### Create R table 1
```{r}
## Rtable_1 shows the demographics stats
Rtable_1 = 
  table1(~ factor(Sex) + Age| Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_1
```


```{r}
## Rtable_2 shows the physiology status factor stats
Rtable_2 = 
  table1(~ factor(HighBP) + factor(HighChol) + BMI + factor(weight_status) | Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_2
```


```{r}
## Rtable_3 shows the social-economic factor stats
Rtable_3 = 
  table1(~ factor(Education) + factor(Income) | Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_3
```

```{r}
## Rtable_4 shows the physical-habits factor stats
Rtable_4 = 
  table1(~ factor(Smoker) + factor(PhysActivity) + factor(Fruits) + factor(Veggies) + factor(HvyAlcoholConsump) + factor(AnyHealthcare) | Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_4
```

```{r}
## Rtable_5 shows the symptoms/disease related stats
Rtable_5 = 
  table1(~ factor(Stroke) + factor(HeartDiseaseorAttack) + factor(DiffWalk) | Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_5
```




### Bivariate Analysis

#### High Blood Pressure in Different Diabetes Status

