---
title: "Data cleaning and EDA"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ggplot2)
library(table1)
```

* Data Description: 

The dataset was collected by the Behavioral Risk Factor Surveillance System (BRFSS) in 2015. This original dataset contains responses from 441,455 individuals and has 330 features. For this project, we use the one with 253,680 responses and a three-class diabetes target variable indicating no diabetes, prediabetes, or diabetes. The potential factors for diabetes are biological factors such as sex, blood pressure, cholesterol, BMI, heart disease, social factors such as education and income, and habits of smoking, physical exercise, eating fruit etc..

* `Diabetes_012`: 0 is for no diabetes or only during pregnancy, 1 is for prediabetes, and 2 is for diabetes. 
* `High BP`: 0 = no high BP 1 = high BP
* `HighChol`: 0 = no high cholesterol 1 = high cholesterol
* `Cholcheck`: 0 = no cholesterol check in 5 years 1 = yes cholesterol check in 5 years
* `BMI`: Body Mass Index
* `Smoker`: Have you smoked at least 100 cigarettes in your entire life? [Note: 5 packs = 100 cigarettes] 0 = no 1 = yes
* `Stroke`: (Ever told) you had a stroke. 0 = no 1 = yes
* `HeartDiseaseorAttack`: coronary heart disease (CHD) or myocardial infarction (MI) 0 = no 1 = yes
* `PhysActivity`: physical activity in past 30 days - not including job 0 = no 1 = yes
* `Fruits`: Consume Fruit 1 or more times per day 0 = no 1 = yes

## Read in Data

```{r}
diabetes_raw <- read_csv("data/Diabetes_012_data.csv") |>
  # factor Diabetes to 0, 1, 2
  mutate(Diabetes_012 = factor(Diabetes_012, levels = c(0, 1, 2)),
         HighBP = factor(HighBP, levels = c(0, 1), labels = c("No", "Yes"))) 

head(diabetes_raw)
```

## Exploratory Data Analysis

### Univariate Analysis

####  Diabetes Type

```{r}

diabetes_raw$Diabetes_012 <- factor(diabetes_raw$Diabetes_012, levels = c(0, 1, 2),
                                     labels = c("No Diabetes", "Pre-Diabetes", "Diabetes"))

# Use ggplot2 to plot the frequency chart
ggplot(data = diabetes_raw) +
  geom_bar(mapping = aes(x = Diabetes_012), fill = "skyblue") +
  labs(x = "Diabetes Status", y = "Frequency", title = "Frequency of Diabetes Categories") +
  theme_minimal()
```

####  High Blood Pressure

```{r}
ggplot(data = diabetes_raw) +
  geom_bar(mapping = aes(x = HighBP), fill = "lightgreen") +
  labs(x = "High Blood Pressure", y = "Frequency", title = "Frequency of High Blood Pressure") +
  theme_minimal()
```

#### bmi categorization

```{r}
# write a function to categorize bmi values to diverse weight status
categorize_bmi <- function(bmi) {
  if (bmi < 18.5) {
    return("Underweight")
  } else if (bmi >= 18.5 & bmi < 25) {
    return("Normal weight")
  } else if (bmi >= 25 & bmi < 30) {
    return("Overweight")
  } else {
    return("Obesity")
  }
}
# add a column named weight_status to the data set
diabetes_raw$weight_status <- sapply(diabetes_raw$BMI, categorize_bmi)

```

#### Plot: Association between Diabetes Status and blood pressure status
```{r}
ggplot(diabetes_raw, aes(x = Diabetes_012, fill = HighBP)) +
  geom_bar(position = "fill") +
  ylab("Blood Pressure Status") +
  xlab("Diabetes Status") +
  ggtitle("Association between Diabetes Status and Blood Pressure Status") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal()
```

#### Plot: Association between Diabetes Status and Weight Status
```{r}
ggplot(diabetes_raw, aes(x = Diabetes_012, fill = weight_status)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  xlab("Diabetes Status") +
  ggtitle("Association between Diabetes Status and Weight Status") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal()
```

#### Plot: Box plot distribution of BMI in different Diabetes status
```{r}
plotBMI = ggplot(diabetes_raw, aes(x = BMI, y = Diabetes_012, fill = Diabetes_012))+
        geom_boxplot(alpha = 0.5) +
        theme(legend.position = 'none') +
        scale_fill_brewer(palette = 'Pastel1') +
        ylab("Diabetes Status") +
        xlab("BMI Distribution") +
        ggtitle("Box Plot Distribution of BMI over Diabetes Status") +
        theme_minimal()
        
  
  
plotBMI
```

#### Plot: Smoking Status over Diabetes Status
```{r}
plot_smoke = 
  ggplot(diabetes_raw, aes(Smoker, ..count..)) + 
  geom_bar(aes(fill = Diabetes_012), position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Smoking Status") +
  ggtitle("Smoking Status Frequencies over Diabetes Status") +
  theme_minimal()

plot_smoke
```
### Plot: Fruits or Veggies over different Diabetes Status
```{r}
mosaicplot(Fruits~Veggies,data=diabetes_raw,col=c("Light Pink","Light Blue"))
```



#### Plot: Distribution of Education Level over different Diabetes Status
```{r}
plot_education = 
  ggplot(diabetes_raw, aes(x = Education, ..count..)) + 
  geom_bar(aes(fill = Diabetes_012), position = 'dodge') +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Education Level") +
  ggtitle("Education Level Frequencies over Diabetes Status") +
  theme_minimal()

plot_education
```

#### Plot: Distribution of Income Level over different Diabetes Status
```{r}
plot_income = 
  ggplot(diabetes_raw, aes(Income, ..count..)) + 
  geom_bar(aes(fill = Diabetes_012), position = "dodge") +
  scale_fill_brewer(palette = 'Pastel1') +
  ylab("Frequencies") +
  xlab("Income Level") +
  ggtitle("Income Level Frequencies over Diabetes Status") +
  theme_minimal()

plot_income
```

#### Statistical Test 1.1 P-value function
```{r}
pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a one-side anova test
        p <- aov(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}
```


#### Create R table 1
```{r}
## Rtable_1 shows the demographics stats
Rtable_1 = 
  table1(~ factor(Sex) + Age| Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_1
```


```{r}
## Rtable_2 shows the physiology status factor stats
Rtable_2 = 
  table1(~ factor(HighBP) + factor(HighChol) + BMI + factor(weight_status) | Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_2
```


```{r}
## Rtable_3 shows the social-economic factor stats
Rtable_3 = 
  table1(~ factor(Education) + factor(Income) | Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_3
```

```{r}
## Rtable_4 shows the physical-habits factor stats
Rtable_4 = 
  table1(~ factor(Smoker) + factor(PhysActivity) + factor(Fruits) + factor(Veggies) + factor(HvyAlcoholConsump) + factor(AnyHealthcare) | Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_4
```

```{r}
## Rtable_5 shows the symptoms/disease related stats
Rtable_5 = 
  table1(~ factor(Stroke) + factor(HeartDiseaseorAttack) + factor(DiffWalk) | Diabetes_012, data=diabetes_raw, overall=F, extra.col=list(`P-value`=pvalue))

Rtable_5
```




### Bivariate Analysis

#### High Blood Pressure in Different Diabetes Status

