---
title: "Statistical Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ggplot2)
library(reshape2)
library(infotheo)
```


## Feature Selection

To investigate the relationship between the variables High Blood Pressure, High Cholesterol Level, BMI, Sex, Fruits Consumption, Veggies Consumption, Smoking, Income Level, Physical Acitivity, and diabetes status, we conducted correlation and ANOVA statistical tests and build a predictive model. 

First, we read in the dataset. 

```{r read data01,echo=FALSE}
diabetes_data01 <- read_csv("data/Diabetes_012_data.csv", show_col_types = FALSE)
diabetes_data01 <- 
  diabetes_data01 %>%
  dplyr::select(Diabetes_012, HighBP, HighChol, BMI, Smoker, 
                Income, Sex, PhysActivity, Fruits, Veggies) %>%
  filter(Diabetes_012 == '2' | Diabetes_012 == '0') %>%
  mutate(Diabetes_012 = ifelse(Diabetes_012 == "2", "1", Diabetes_012)) %>%
  mutate(Diabetes_012 = as.numeric(Diabetes_012))
```


## Correlation analysis for Continuous Variables
```{r correlation}
correlationX <- 
  diabetes_data01 |>
  dplyr::select(-starts_with("Diabetes_012"))

correlations <- cor(correlationX,diabetes_data01$Diabetes_012)
cor_df <- data.frame(Variable = names(correlationX), Correlation = correlations)
ggplot(cor_df, aes(x = Variable, y = Correlation)) +
  geom_bar(stat = "identity", fill = "purple") +
  theme_minimal() +
  labs(title = "Correlation with Diabetes_012", x = "Variable", y = "Correlation") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

High Blood Pressure, High Cholesterol, and BMI show a positive correlation, indicating that an increase in these factors is associated with higher diabetes metrics. 

Physical Activity and increased Income exhibit a slight negative correlation, suggesting they may contribute to lower diabetes measures. 

Fruits, sex, smoker and vegetable intake also show varying degrees of correlation, reflecting the complex interplay between socioeconomic factors, dietary habits, and health outcomes.



## ANOVA test
```{r anova}
dependent_var <- diabetes_data01[[1]]
independent_vars <- diabetes_data01[,-1]
independent_vars <- data.frame(lapply(independent_vars, function(x) as.factor(x)))
lm_model <- lm(dependent_var ~ ., data = independent_vars)

# Perform ANOVA
anova_results <- anova(lm_model)

# Display the ANOVA results
anova_results|>
  knitr::kable()
```

We can see that all variables have significant small p-values.

## Chi-Square test
```{r chisq}
chi_square_results <- data.frame(Variable = character(), ChiSquare = numeric(), PValue = numeric(), stringsAsFactors = FALSE)

# Perform chi-square test for each independent variable
for (var in names(independent_vars)) {
  # Create a contingency table
  table <- table(dependent_var, independent_vars[[var]])
  
  # Perform chi-square test
  test <- chisq.test(table)
  
  # Store the results
  chi_square_results <- rbind(chi_square_results, data.frame(Variable = var, ChiSquare = test$statistic, PValue = test$p.value))
}

# Order results by p-value
chi_square_results <- chi_square_results %>%
  arrange(PValue) 

# Print the results
chi_square_results|>
  knitr::kable()
```

We can see that all variables have significant p-values around to 0.
